---
title: "Assignment 1: Introduction"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on introductory material. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document (marked with >).
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After completing your assignment, fill out the assignment completion survey in Sakai. 

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-01-18

## Course Setup

1. Post the link to your forked GitHub repository below. Your repo should include one or more commits and an edited README file.  

>  Link: https://github.com/aislinnmcl/Water_Data_Analytics_2022

## Data Visualization Exercises

2. Set up your work session. Check your working directory, load packages `tidyverse`,  `dataRetrieval`, and `zoo`. Set your ggplot theme as theme_classic (you may need to look up how to set your theme).

```{r, message = FALSE}

# get working directory
getwd()

# load packages
library(tidyverse)
library(dataRetrieval)
library(zoo)

# set theme
mytheme <-
  theme_classic()

theme_set(mytheme)

```

3. Upload discharge data for the Eno River at site 02096500 for the same dates as we studied in class (2012-01-01 through 2021-12-31). Obtain data for discharge. Rename the columns with informative titles, as we did in class. 

```{r}

# import data 
eno_discharge <- readNWISdv(siteNumbers = "02096500",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2012-01-01",
                     endDate = "2021-12-31")

# rename columns
eno_discharge <-
  eno_discharge %>%
  rename(discharge_cfs = X_00060_00003, approval_code = X_00060_00003_cd)

```


4. Build a plot called EnoPlot2. Use the base plot we made in class and make the following changes: 

* Add a column to your data frame for discharge in meters cubed per second. hint: package dplyr in tidyverse includes a `mutate` function
* Add a column in your data frame for a 30-day rolling mean of the metric discharge. (hint: package dplyr in tidyverse includes a `mutate` function. hint: package zoo includes a `rollmean` function)
* Create two geom_line aesthetics, one for daily discharge (meters cubed per second) and one for rolling mean of discharge. Color these differently.
* Update your ggplot theme. I suggest "classic." (hint: https://ggplot2.tidyverse.org/reference/ggtheme.html)
* Update axis names 
* Change the y axis from a linear to a log10 axis (hint: google "ggplot logged axis")
* Add a legend. (hint: Google "add legend two geom layers ggplot")

```{r}  

# add columns for cubic meters per second and 30-day rolling mean
eno_discharge <-
  eno_discharge %>%
  mutate(discharge_m3s = discharge_cfs * 0.028316847) %>%
  mutate(mean_30d = rollmean(discharge_m3s, 30, fill = NA))

# plot eno discharge data
EnoPlot2 <-
  ggplot(eno_discharge) +
  geom_line(aes(x = Date, y = discharge_m3s, color = "Daily Value"), alpha = 0.7) +
  geom_line(aes(x = Date, y = mean_30d, color = "Rolling Mean")) +
  labs(x = "Year", y = "Discharge (m^3 per second)") +
  scale_y_continuous(trans='log10') +
  scale_color_manual(name = "", values = c("Daily Value" = "goldenrod2", "Rolling Mean" = "steelblue"))

EnoPlot2
```

5. In what ways was the second plot a more effective visualization than the first? 

> ANSWER: The log transformation on the y-axis allows you to view the variation more clearly because the data aren't scrunched up against the x-axis. Also the rolling mean shows a smoother picture of the peaks/troughs in the measurements.

6. What portions of the coding were challenging for you?

> ANSWER: Mostly refreshing my memory on commands - I've covered a decent amount of this material (incl USGS package) in EDA and coding for my MP. It's always a little tricky to get plots to look exactly how I want them (e.g. good colors, legend that shows labels/data I want)

7. Interpret the graph you made. What are the things you notice about within- and across-year variability, as well as the differences between daily values and 30-day rolling mean?

> ANSWER: There seems to be some seasonality within years - regular points at which the discharge dips and peaks. Across years, the max discharge in m^3/s seems to be increasing as does the min discharge in m^3/s - the annual max daily value seems to generally have increased YoY since 2012 and the annual min daily value seems to have been increasing since 2016. The rolling mean helps to visually emphasize the discharge seasonality and other longer term trends while the daily values highlight the variation in annual max/min daily values.

